sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-03 doc/paper-jss/owls.Rmd
Streamline chunk settings
Added figure caption in the owl appendix

```{r setup-owls, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.7,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dev = "png",
  dpi = 150,
  optipng = "-strip all"
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

@@

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations. The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`). This mesh uses a cutoff distance of 1.5 units (or 150 km).

```{r mesh-owls, echo=TRUE, out.width="4in"}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)

@@

We can inspect randomized quantile residuals, although this method has known issues [@thygesen2017a].

```{r qqnorm-owls, out.width="60%"}
qqnorm(pred$resid)
qqline(pred$resid)
```

@@

```{r theme-owls, echo=FALSE}
theme_set(theme_bw() + theme(
  plot.title = element_text(size = 12),
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank() # lat/lon is clear from axis text
)) 
```

@@

The main effect of NAO indicates that there is a weak average positive effect of annual mean NAO on overall counts `r round(b[b$term == "nao", 2], 2)` (`b[b$term == "nao", 2]`), while for individual count locations the median estimated effects of NAO range from `r round(min(p_mean$nao_sim), 2)` to `r round(max(p_mean$nao_sim),2)` (`range(p_mean$nao_sim)`).
Based on the 80% CI calculated above, the resulting minimum estimated effects of NAO for each count location range from `r round(min(p_mean$nao_lwr), 2)` to `r round(max(p_mean$nao_lwr),2)` (`range(p_mean$nao_lwr)`) and the maximums range from `r round(min(p_mean$nao_upr), 2)` to `r round(max(p_mean$nao_upr),2)` (`range(p_mean$nao_upr)`). 
All these effects are in link space (`link = "log"`) and could be exponentiated to give the expected multiplicative contribution to the predicted owl counts; however, the additive nature of the effect in link space is more intuitive to interpret visually, so this is what we will plot on our maps:

```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="90%", echo=FALSE}
ggplot(data = p_proj) +
  geom_point(
    data = p_mean, aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none") +
  labs(colour = "Estimated effect in log space") +
  ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
  theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
```

```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%",fig.asp=0.3}
p1 <- ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none", colour = "none") +
  # labs(colour = "Lower 90% CI") +
  ggtitle("Lower 80% CI") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())

@@

These are the predictions of winter owl counts that incorporate all fixed effects and random effects. 
The mean estimated count is `r round(mean(exp(p$est)), 2)`, which is essentially equal to the observed mean of `r round(mean(snow$count), 2)`.
The maximum estimated count was `r round(max(exp(p$est)))`, but the colour axis on this plot has been transformed and trimmed to improve the visibility of variation in the lower range of count estimates.

```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

We can look at the spatial random effects (`omega_s`) that represent consistent deviations in space that are not accounted for by our fixed effects. 
In other words, these deviations represent consistent biotic and abiotic factors that are affecting winter owl counts.
These could be landcover, proximity to the breeding grounds, and effects of flyways and barriers such as coasts.

```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

Finally, we can look at the spatiotemporal random effects (`epsilon_st`) that represent deviations from the fixed effect predictions and the spatial random effect deviations. 
These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
For owls, this is most likely to be factors like landcover changes, prey abundance cycles, and snow depth.

```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

In this case, these only include the overall effect of mean annual NAO and the random intercepts for year.

```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

```{r setup-owls, include = FALSE, cache=FALSE}
theme_set(theme_bw() + theme(
  plot.title = element_text(size = 12),
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank() # lat/lon is clear from axis text
)) 
```

@@

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations (Fig.\ \ref{fig:mesh-owls}). The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`). This mesh uses a cutoff distance of 1.5 units (or 150 km).

```{r mesh-owls, echo=TRUE, out.width="4in", fig.cap="Mesh used for model is based on X and Y that use an Albers projection divided by 100000 to give units of 100 km and a cutoff distance of 1.5 units (or 150 km).", fig.pos='h', fig.align='center'}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)

@@

We can inspect randomized quantile residuals (Fig.\ \ref{fig:qqnorm-owls}), although this method has known issues [@thygesen2017a].

```{r qqnorm-owls, out.width="60%", fig.cap="Randomized quantile residuals.", fig.pos='h', fig.align='center'}
qqnorm(pred$resid)
qqline(pred$resid)
```

@@

@@

The main effect of NAO indicates that there is a weak average positive effect of annual mean NAO on overall counts `r round(b[b$term == "nao", 2], 2)` (`b[b$term == "nao", 2]`), while for individual count locations the median estimated effects of NAO range from `r round(min(p_mean$nao_sim), 2)` to `r round(max(p_mean$nao_sim),2)` (`range(p_mean$nao_sim)`).
Based on the 80% CI calculated above, the resulting minimum estimated effects of NAO for each count location range from `r round(min(p_mean$nao_lwr), 2)` to `r round(max(p_mean$nao_lwr),2)` (`range(p_mean$nao_lwr)`) and the maximums range from `r round(min(p_mean$nao_upr), 2)` to `r round(max(p_mean$nao_upr),2)` (`range(p_mean$nao_upr)`). 
All these effects are in link space (`link = "log"`) and could be exponentiated to give the expected multiplicative contribution to the predicted owl counts; however, the additive nature of the effect in link space is more intuitive to interpret visually, so this is what we will plot on our maps (Fig.\ \ref{fig:plot-map-ci}).

```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="90%"}
ggplot(data = p_proj) +
  geom_point(
    data = p_mean, aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none") +
  labs(colour = "Estimated effect in log space") +
  # ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
  theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
```

```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%",fig.asp=0.3, fig.cap="Map of combined main and spatially varying effects of NAO on Snowy Owl count with confidence intervals. In the west, the lower values span zero implying a no evidence of an effect at this confidence level, while in the southeast the lowest estimates are still positive. Note that the values are in log space and thus additive. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).", fig.pos='h', fig.align='center'}
p1 <- ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none", colour = "none") +
  # labs(colour = "Lower 90% CI") +
  ggtitle("Lower 80% CI") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())

@@

For each year in the model, we can generate predictions of winter owl counts that incorporate all fixed effects and random effects (Fig.\ \ref{fig:plot-all-effects-owls}). 
The mean estimated count is `r round(mean(exp(p$est)), 2)`, which is essentially equal to the observed mean of `r round(mean(snow$count), 2)`.
The maximum estimated count was `r round(max(exp(p$est)))`, but the colour axis on this plot has been transformed and trimmed to improve the visibility of variation in the lower range of count estimates.

```{r plot-all-effects-owls, echo=FALSE, eval=TRUE, dpi=220, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='h', fig.align='center'}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

We can look at the spatial random effects (`omega_s`) that represent consistent deviations in space that are not accounted for by our fixed effects (Fig.\ \ref{fig:plot-spatial-effects-owls}). 
In other words, these deviations represent consistent biotic and abiotic factors that are affecting winter owl counts.
These could be landcover, proximity to the breeding grounds, and effects of flyways and barriers such as coasts.

```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200, fig.cap="Spatial random effects that apply across all years.", fig.pos='h', fig.align='center'}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

We can also look at the spatiotemporal random effects (`epsilon_st`) that represent deviations from the fixed effect predictions and the spatial random effect deviations (Fig.\ \ref{fig:plot-spatiotemporal-effects-owls}). 
These represent biotic and abiotic factors causing spatial correlation that are changing through time and are not accounted for in the model. 
For owls, this is most likely to be factors like landcover changes, prey abundance cycles, and snow depth.

```{r plot-spatiotemporal-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="Spatiotemporal random effects.", fig.pos='h', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

Finally, the non-random field based elements in the model are represented by `est_non_rf` variable in the prediction dataframe, and in this case only include the overall effect of mean annual NAO and the random intercepts for year (Fig.\ \ref{fig:plot-fixed-effects-owls}).

```{r plot-fixed-effects-owls, echo=FALSE, eval=TRUE, dpi=210, fig.cap="The facet colours represent the non-random field based elements in this model.", fig.pos='h', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

### 2022-11-03 doc/paper-jss/paper-jss.Rmd
Added figure caption in the owl appendix

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  cache.comments = FALSE,
  dev = "png",
  dpi = 180
  # optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "> ")
)
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

>>>>

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = FALSE,
  dev = "png",
  dpi = 180
  # optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "> ")
)
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```
